---
title: "London Homicides 2008 to 2020"
author: "David Harrington"
output_dir: docs
output: 
  flexdashboard::flex_dashboard:
    social: menu
    source_code: embed
---

```{r global, include=FALSE}

library(flexdashboard)
library(knitr)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(lubridate)
library(leaflet)
library(leaflet.extras)
library(tidyverse)
library(reshape)
library(ggridges)
library(rgeos)
library(ggmap)
library(maptools)
library(broom)
library(mapproj)
library(htmlwidgets)
library(htmltools)
library(plotly)

    # data import
    
    library(readxl)
    mmap18 <- read_excel("C:/Users/david/OneDrive - RMIT University/Data Visualisation and Communication/London Homicides Project/mmap.xlsx")
    
    
    # clean up the orginal data set and eliminate redundant variables
    
    library(dplyr)
    mmap18 = select(mmap18, -6,-7,-8,-13,-14,-16,-17,-18,-19)
    
    # reorder columns
    mmap18 <- mmap18[ , c(2, 1, 8, 9,5,4,3,10,6,11,7)]  
    
    
    # update column names
    colnames(mmap18) <- c("date", "ID", "latitude", "longitude", "ethnicity", "sex", "age", "ageGroup", "weapon", "borough", "status")
    
    # change strings to factors
    mmap18 <- mmap18 %>% mutate_if(is.character, as.factor) 
    
    # import the 2019 murder map data into R
    mmap19 <- read_excel("C:/Users/david/OneDrive - RMIT University/Data Visualisation and Communication/London Homicides Project/Murdermap2019.xlsx")
    
    mmap19 = select(mmap19, -3,-6)
    
    
    mmap19$longitude <- as.numeric(mmap19$longitude)
    
    
    # change strings to factors
    mmap19 <- mmap19 %>% mutate_if(is.character, as.factor) 
    
    colnames(mmap19) <- c("date", "ID", "latitude", "longitude", "ethnicity", "sex", "age", "ageGroup", "weapon", "borough", "status")
    
    # import the 2019 murder map data into R
    mmap20 <- read_excel("C:/Users/david/OneDrive - RMIT University/Data Visualisation and Communication/London Homicides Project/Murdermap2020.xlsx")
    
    mmap20 = select(mmap20, -3,-6)
    
    mmap20 <- mmap20 %>% mutate_if(is.character, as.factor) 
    
    colnames(mmap20) <- c("date", "ID", "latitude", "longitude", "ethnicity", "sex", "age", "ageGroup", "weapon", "borough", "status")
    
    
    ##Merge mmap19 and mmap 20 using rbind function
    mmap19_20 = rbind(mmap19,mmap20)
    
    #change the levels for "sex" to align with mmap18
    mmap19_20 <- mmap19_20 %>%
      mutate(sex = fct_recode(sex,
                              "M" = "Male",
                              "F" = "Female")) 
    
    # merge mmap19_20 with mmap18 to create mmap (integrated cases from 2008 to 2020)
    mmap = rbind(mmap18, mmap19_20)
    
    
    # See all levels in weapon variable
    levels(mmap$weapon) 
    
    # Replicate weapon column to enable re-factoring
    mmap2 = cbind(mmap, weapon2=rep(mmap$weapon)) 
    
    
    # Change levels in Weapon column to include just 4 levels Knife, Gun, Assault and Other
    mmap3 <- mmap2 %>% mutate( weapon = fct_recode(weapon,
                                                   "Other" = "Arson",
                                                   "Other" = "Vehicle",
                                                   "Other" = "Drug",
                                                   "Other" = "Poison",
                                                   "Other" = "Unknown",
                                                   "Other" = "Ligature",
                                                   "Other" = "Strangulation",
                                                   "Other" = "Fire",
                                                   "Other" = "Other",
                                                   "Other" = "Blunt Object",
                                                   "Assault" = "None"))
    
    
    
    # clean up the factor names within borough variable to enable the bind with boroughpop (ensure they are exactly the same!)
    mmap33 <- mmap3 %>% mutate(borough = fct_recode(borough, 
                                                    "Westminster" = "City of Westminster",
                                                    "Kingston upon Thames" = "Kingston-upon-Thames",
                                                    "Kingston upon Thames" = "Kingston Upon Thames",
                                                    "Kingston upon Thames" = "Kingston-Upon-Thames",
                                                    "Richmond upon Thames" = "Richmond"))
    
    
    # Replicate date column
    mmap4 = cbind(mmap33, date2=rep(mmap3$date))
    
    
    # create new data frame mmap5 and separate date variable from mmap4 into year, month and day and save them as integers
    mmap5 = separate(mmap4, date2, c("year", "month", "day"), "-") %>%
      mutate_if(is.character, as.integer) 
    
    mmap5 <- mmap5 %>% mutate(status2 = fct_recode(status, 
                                                   "Awaiting Trial" = "Awaiting trial",
                                                   "Awaiting Trial" = "Awaiting Outcome",
                                                   "Solved" = "Solved",
                                                   "Solved" = "Self Defence",
                                                   "Unsolved" = "Suspect dead",
                                                   "Unsolved" = "Unsolved"))
    
    
    
    mmap5 <- mmap5 %>% mutate(ethnicity = fct_recode(ethnicity, 
                                                     "Other" = "Mixed",
                                                     "Black" = "Black",
                                                     "Black" = "Black or Black British",
                                                     "White" = "White",
                                                     "White" = "White or White British",
                                                     "Asian" = "Asian",
                                                     "Asian" = "Asian or Asian British",
                                                     "Other" = "Unknown",
                                                     "Other" = "NA",
                                                     "Other" = "Any Other Ethnic Appearance"))
    
    levels(mmap5$ethnicity)
    
    # read the boroughpop file into R to access borough population data
    boroughpop <- read_excel("boroughpop.xlsx")
    
    
    # change column names in boroughpop
    colnames(boroughpop) = c("borough", "population")
    
    # join mmap5 and boroughpop data frames using "borough" 
    mmap6 <- left_join(mmap5, boroughpop, by = "borough")
    
    
    # make bourough a factor variable
    mmap6$borough <- as.factor(mmap6$borough)
    
    
    # create new variable for counting homicides by borough and homicides per 100,000 population by borough
    mmap6_sum <- mmap6 %>% select(borough, population) %>% group_by(borough) %>% mutate(count = n(), population = population, homicide_rate = (count*100000)/11/population) %>% distinct()
    
    # create mmap 7 by filtering mmap5 to just include full year data to end of calendar 2019
    mmap7 <- mmap6 %>% filter(year != "2020")
    
    
    mmap7 <-data.frame(mmap7)
    
    # remove NA cases for ethicity
    mmap8 <- mmap7 %>% filter(ethnicity != "NA")
    
    # save mmap6_sum as data frame and name the data frame mmap9                                                                               
    mmap9 <- as.data.frame(mmap6_sum)
    
    # create new data frame mmap10 with just borough and homicide rate variables
    mmap10 <- mmap9 %>% select(borough, homicide_rate)
    
    # round the homicide_rate variable to 2 decimal places
    mmap10$homicide_rate <- round(mmap10$homicide_rate, 2)
    
    # arrange the data frame based on homicide rate variable descending order
    mmap10 <- mmap10 %>% arrange(desc(mmap10$homicide_rate)) 
    
    
    mmap7$year <- as.factor(mmap7$year)   

    # calculate sum borough population based on the 2016 ONS data for boroughs
GreaterLondonPop <- sum(boroughpop$population)
GreaterLondonPop

# create vector for Greater London Population for 2008 to 2019
 GreaterLondonPopMillions <- c(7.8, 7.9, 8.0, 8.15, 8.25, 8.35, 8.5, 8.7, 8.82, 8.85, 8.9, 9.0)
 
# Greater London Population in 100,000s
GreaterLondon <- GreaterLondonPopMillions*10

# Greater London total homicides

LondonHomicides <- c(155, 130, 128, 121, 103, 113, 90, 120, 107, 130, 131, 149)

Year <- c(2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019)

# merge vectors into a data frame

London <- data.frame(Year, LondonHomicides, GreaterLondon)

London


# Number of homicides per 100,000 population, 2008-2019

LondonHomicideRate <- London %>% mutate(Homicide_Rate = LondonHomicides/GreaterLondon)

LondonHomicideRate$Homicide_Rate <- round(LondonHomicideRate$Homicide_Rate,2)

LondonHomicideRate$Year <- as.factor(LondonHomicideRate$Year)

str(LondonHomicideRate)

sex <- table(mmap5$sex) # Victim gender table
addmargins(sex) # Plus totals

# create new data frame to show ageGroup data by year
ageyear <- mmap7 %>% 
  select(ageGroup, year) %>%
  group_by(ageGroup, year) %>%
  tally() %>%
  spread(ageGroup, n) 


# calculate total homicides per ageGroup across 2008 to 2019
homicidesA <- sum(ageyear$`A. Child 0-6`, na.rm= TRUE)
homicidesB <- sum(ageyear$`B. Child 7-12`, na.rm= TRUE)
homicidesC <- sum(ageyear$`C. Teen 13-16`, na.rm= TRUE)
homicidesD <- sum(ageyear$`D. Teen 17-19`, na.rm= TRUE)
homicidesE <- sum(ageyear$`E. Adult 20-24`, na.rm= TRUE)
homicidesF <- sum(ageyear$`F. Adult 25-34`, na.rm= TRUE)
homicidesG <- sum(ageyear$`G. Adult 35-44`, na.rm= TRUE)
homicidesH <- sum(ageyear$`H. Adult 45-54`, na.rm= TRUE)
homicidesI <- sum(ageyear$`I. Adult 55-64`, na.rm= TRUE)
homicidesJ <- sum(ageyear$`J. Adult 65 over`, na.rm= TRUE)

# consolidate total homicides per ageGroup across 2008 to 2019 into a new vector homicides_ageGroup
homicides_ageGroup <- c(homicidesA, homicidesB, homicidesC, homicidesD, homicidesE, homicidesF, homicidesG, homicidesH, homicidesI, homicidesJ)

# create vector agegp with unique values for the variable "ageGroup" in mmap5
agegp <- unique(mmap5$ageGroup)

# refactor to order the values for the variable "ageGroup" from youngest to oldest
agegp <- factor(c("A. Child 0-6", "B. Child 7-12", "C. Teen 13-16", "D. Teen 17-19", "E. Adult 20-24", "F. Adult 25-34","G. Adult 35-44","H. Adult 45-54","I. Adult 55-64","J. Adult 65 over"))

# create vector to calculate average homicides per year over 11 years from 2008 to 2019
homicides_ageGroup_avg <- homicides_ageGroup/11

# create population vector by age group - Source: ONS small area population estimates mid-2016 (latest ONS survey for UK)
popn <- c(872040, 659384, 374989, 278134, 557975, 1654605, 1408946, 1151408, 828359, 1039161) 

# create homicide-rate2 vector
homicide_rate2  <- homicides_ageGroup_avg / popn * 100000 


# Create a data frame from vectors
mmap11 <- data.frame(agegp, round(homicide_rate2,2)) 

colnames(mmap11) <- c("Age_Group", "Homicide_Rate")

# calculate mean and median age for all homicides in London 2008-2020
meanAge <- mean(mmap5$age)
meanAge

medianAge <- median(mmap5$age)
medianAge

# Trends in Status

## create tables showing probability of being solved given weapon, ethnicity and borough

# by weapon type
status <- table(mmap5$weapon, mmap5$status, dnn = c("weapon", "status"))
status

statusProp <- prop.table(status, 1)
statusProp

# save as data frame
weaponStatus <- data.frame(statusProp)

# filter to just include probability of being solved
weaponStatus1 <- weaponStatus %>% filter(status == "Solved")

# change to percent
weaponStatus1$percent <- round(weaponStatus1$Freq,2)*100

mmap13 <- weaponStatus1 %>% arrange(weaponStatus1$percent)

mmap13$percent <- as.integer(mmap13$percent)

p1 <- ggplot(mmap7, aes(x = year)) +
  geom_bar(stat='count', fill = "navyblue", color = "white") +
  geom_label(stat = "count", aes(label = ..count..,y = ..count..), size = 3) +
  labs(x = "Year", y = "Number of Homicides", title = "London homicides, 2008-2019")+
  theme_bw()

p3 <- ggplot(LondonHomicideRate, aes(x= Year, y=Homicide_Rate)) +
geom_bar(stat="identity", fill="navyblue", color = "white")+
  geom_label(aes(label = Homicide_Rate, vjust = 1),size = 3)+
  labs(x = "Year", y = "Rate", title = "London homicide rate per 100,000")+
  theme_bw()

# by ethnicity

levels(mmap8$ethnicity)

status2 <- table(mmap8$ethnicity, mmap8$status, dnn = c("ethnicity", "status"))
status2

status2Prop <- round(prop.table(status2, 1),2)
status2Prop

# save as data frame
ethnicityStatus <- data.frame(status2Prop)


# filter to just include probability of being solved

ethnicityStatus1 <- ethnicityStatus %>% filter(status == "Solved")

# change to percent
ethnicityStatus1$percent <- round(ethnicityStatus1$Freq,2)*100

mmap14 <- ethnicityStatus1 %>% arrange(ethnicityStatus1$percent)

# by ageGroup

levels(mmap8$ageGroup)

status4 <- table(mmap8$ageGroup, mmap8$status, dnn = c("ageGroup", "status"))
status4

status4Prop <- round(prop.table(status4, 1),2)
status4Prop

# save as data frame
AgeGroupStatus <- data.frame(status4Prop)


# filter to just include probability of being solved

AgeGroupStatus1 <- AgeGroupStatus %>% filter(status == "Solved")

# change to percent
AgeGroupStatus1$percent <- round(AgeGroupStatus1$Freq,2)*100


# by location
status3 <- table(mmap5$borough, mmap5$status, dnn = c("borough", "status"))
status3

status3Prop <- prop.table(status3, 1)
status3Prop

# save as data frame
boroughStatus <- data.frame(status3Prop)

# filter to just include probability of being solved

boroughStatus1 <- boroughStatus %>% filter(status == "Solved")

# change to percent
boroughStatus1$percent <- round(boroughStatus1$Freq,2)*100


# Install ggplot2 mapping packages

library(rgeos)
library(ggmap)
library(maptools)
library(broom)
library(mapproj)


# create Leaflet interactive cloropeth chart for Homicide Rate

# read shp file for just London boroughs
London_Ward2 <- readShapeSpatial("London_Borough_Excluding_MHW.shp")

# convert coordinates into lat long degrees
proj4string(London_Ward2) <- CRS("+init=epsg:27700")
London_Ward2.wgs84 <- spTransform(London_Ward2, CRS("+init=epsg:4326"))


library(leaflet)

# For Leaflet choropleth maps, we need to use a SpatialPolygonDataFrame

class(London_Ward2.wgs84)
names(London_Ward2.wgs84)

# Rename the borough variable in mmap10 to "NAME" to match the "SpatialPolygonsDataFrame" file London_Ward2.wgs84

names(mmap10)[names(mmap10) == "borough"] <- "NAME"

str(mmap10)

mmap10 <- mmap10 %>% mutate(NAME = fct_recode(NAME, 
                                                        
                             "City of London" = "City and County of the City of London",
                            
                           "Westminster" = "City of Westminster"))  

levels(mmap10$NAME)

# merge mmap data into London_Ward.wgs84 SpatialPolygonsDataFrame using "DISTRICT"
merge.mmap10 <- sp::merge(London_Ward2.wgs84, mmap10, by = "NAME", duplicateGeoms = TRUE)


# create a discrete colour scale based on quantiles for five levels

bins <- quantile(
  mmap10$homicide_rate,
  probs = seq(0,1,.2), names = FALSE, na.rm = TRUE)
bins



```
Homicide Summary {data-orientation=Column}
=======================================================================

Column {data-width=600}
-----------------------------------------------------------------------

### The median age for all London homicide victims is 30 (2008 to 2020)

```{r}
p6 <- ggplot(mmap5, aes(x = age))+ 
  geom_density(fill = "red", alpha = 0.5)+ 
  geom_histogram(colour = "white", fill = "navyblue", aes(age, ..density..), alpha = 0.5, bins = 100)+
  labs(x = "Victim Age", y = "Proportion of Homicides")+
  geom_vline(xintercept= median(mmap7$age)) +
  annotate("text",label = "Mean = 34.6",x = 45, y = 0.04, colour = "red") +
  geom_vline(xintercept= mean(mmap7$age),linetype=2) +
  annotate("text",label = "Median = 30.0",x = 41, y = 0.035, colour = "red")+
  scale_x_continuous(limits = c(0, 100))+
   theme_bw()
ggplotly(p6)
```

### After decling from 2008, the number of homicides in London increased after 2014

```{r}

p1 <- ggplot(mmap7, aes(x = year)) +
  geom_bar(stat='count', fill = "navyblue", color = "white") +
  geom_label(stat = "count", aes(label = ..count..,y = ..count..), size = 3) +
  labs(x = "Year", y = "Number of Homicides")+
  theme_bw()
ggplotly(p1)
```

```
```

Column {data-width=500}
-----------------------------------------------------------------------

### Homicide rates are highest in the centre of London - peaking at over 7 victims per 100,000 in City of London (average for 2008 to 2019)

```{r}
  pal <- colorBin(
    "YlOrRd",
    domain = mmap10$homicide_rate,
    bins = bins
  )

# add the colour scale to the choropleth map.
labels <- sprintf(
  "%s
%g homicide_rate",
  merge.mmap10$NAME,
  merge.mmap10$homicide_rate
) %>% lapply(htmltools::HTML)

title <- tags$div(
  HTML('<h3>London Homicide Rate by Borough</h3>')
)

p3 <- leaflet(merge.mmap10) %>%
  setView(lng = -0.118092 , lat = 51.509865, zoom = 10)
p3 %>% addPolygons(
  fillColor = ~pal(homicide_rate),
  weight = 2,
  opacity = 0.5,
  color = "black",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#0000FF",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = labels,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>%
  addLegend(pal = pal,
            values = ~homicide_rate,
            opacity = 0.7, title = "Homicide Rate",
            position = "bottomright") %>%
  addControl(title, position = "topright") %>%
  setView(lng=-0.1, lat=51.51, zoom=10) %>%
  addProviderTiles("Esri.NatGeoWorldMap")
  
```


Column 
-----------------------------------------------------------------------

### The number of black victims of homicide has grown sharply since 2014 (2008-2019)

```{r}
 p16 <- ggplot(mmap8, aes(x = year, fill = ethnicity)) +
  geom_density(stat='count', alpha = 0.6, colour = "black") +
  scale_x_continuous(breaks = c(2008:2019)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Calendar Year", y = "Number of Homicides", fill = "legend")+
   theme_bw()
ggplotly(p16)

``` 

### Death by knife is the largest and most rapdily growing cause of death (2008-2019)

```{r}
 p17 <- ggplot(mmap8, aes(x = year, fill = weapon)) +
  geom_density(stat='count', alpha = 0.6, colour = "black") +
  scale_x_continuous(breaks = c(2008:2020)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Calendar Year", y = "Number of Homicides", fill = "legend")+
   theme_bw()
ggplotly(p17)

```
```
```
By Gender and Age {data-orientation=Column}
=======================================================================

Column {data-width=550}
-----------------------------------------------------------------------


### A higher proportion of male homicides are young (2008-2020)

```{r}

p9 <- ggplot(mmap5, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.8)+
  scale_fill_brewer(palette = "Set1")+
    labs(fill = "Legend")+
  scale_x_continuous("Victim Age")+
  scale_y_continuous("Proportion of Homicides")+
 theme_bw()
ggplotly(p9)
```

### There is a high concentration of male victims at a younger age with a mean age of 33 (2008-2020)

```{r}

p10 <- ggplot(data = mmap5, aes(x = sex, y = age)) + geom_boxplot(outlier.shape = NA) + geom_jitter(alpha = 1/5) +
  ylab("age of victim") +
  stat_summary(fun.y="mean", colour="red", geom="point",shape = 16, size = 4) +
  theme_minimal()
ggplotly(p10)


```

```
```
Column {data-width=500}
-----------------------------------------------------------------------


### There are larger clusters of female victims near the centre of London (2008 - 2020)

```{r}
femaleVictims <- mmap5 %>%
  filter(sex == 'F')


femaleVictims %>% # Map showing all point data
 leaflet() %>%
  setView(lng =-0.1, lat=51.51, zoom=10) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  addScaleBar %>%
  addCircleMarkers(lng = ~longitude, lat = ~latitude, popup = ~paste0("<b>", ID, "</b>", "<br/>", date, "<br/>", age, "<br/>", sex, "<br/>", weapon, "<br/>", borough), radius = 5, clusterOptions = markerClusterOptions)
``` 
```
```
Column
-----------------------------------------------------------------------

### A higher proportion of black homicides are young (2008-2019)

```{r}

p13 <- ggplot(mmap8, aes(x = age, fill = ethnicity)) +
  geom_density(alpha = 0.6)+
  scale_fill_brewer(palette = "Set1")+
    labs(fill = "Legend")+
  scale_x_continuous("Victim Age")+
  scale_y_continuous("Proportion of Homicides")+
 theme_bw()
ggplotly(p13)
``` 


### 77% of all homicide victims in London are male (2008-2019)


```{r}
 p18 <- ggplot(mmap8, aes(x = year, fill = sex)) +
  geom_density(stat='count', alpha = 0.6, colour = "black") +
  scale_x_continuous(breaks = c(2008:2020)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Calendar Year", y = "Number of Homicides", fill = "legend")+
   theme_bw()
ggplotly(p18)

```


```
```

By Ethnicity and Cause of Death {data-orientation=Column}
=======================================================================

Column {data-width=600}
-----------------------------------------------------------------------


### Black victims of homicide are younger than victims from White, Asian or Other ethnicities (2008-2019)

```{r}
p14 <- ggplot(data = mmap8, aes(x = reorder(ethnicity, age, "mean"), y = age), na.rm = TRUE)+ geom_boxplot(outlier.shape = NA) + geom_jitter(alpha = 1/5) +
  ylab("Age of victim") +
  xlab("Ethnicity")+
  stat_summary(fun="mean", colour="red", geom="point",shape = 16, size = 4) +
  theme_minimal()
ggplotly(p14)

``` 

### Homicides involving black victims are less likely to get solved in the criminal justice system (2008-2020)

```{r}

p23 <- ggplot(mmap14, aes(x= reorder(ethnicity, percent), y=percent))+
geom_bar(stat="identity", fill="navyblue", color = "black")+
  coord_flip()+
  labs(x = "Ethnicity", y = "Percent Solved")+
  theme_minimal()
ggplotly(p23)

```
```
```

Column {data-width=550}
-----------------------------------------------------------------------

### Homicides by victim ethnicity are generally dispersed evenly across London with a higher concentration of Black victims in Eastern boroughs (2008-2020)

```{r}
pal2 <- colorFactor(palette = c("darkgreen", "purple", "blue", "red"),
                    levels = c("Black", "White", "Asian", "Other")) # Creating a 4 colour palette

black <- mmap5 %>%
  filter(ethnicity == "Black") # Grouping layers - black

white <- mmap5 %>%
  filter(ethnicity == "White") # Grouping layers - white

asian <- mmap5 %>%
  filter(ethnicity == "Asian") # Grouping layers - asian

other <- mmap5 %>%
  filter(ethnicity == "Other") # Grouping layers - other



mmap5 %>%
 leaflet() %>%
  setView(lng=-0.1, lat=51.51, zoom=10) %>%
  addProviderTiles("Esri.NatGeoWorldMap") %>%
  addScaleBar(position = "bottomleft") %>%
  addCircleMarkers(data = black,
                   lng = ~longitude, 
                   lat = ~latitude, 
                   radius = 2, 
                   color = ~pal2(ethnicity),
                   group = "Black",
                   popup = ~paste("<b>", ID, "</b>", "<br/>", date, "<br/>", age, "<br/>", sex, "<br/>", weapon, "<br/>", status)) %>%
  addCircleMarkers(data = white,
                   lng = ~longitude, 
                   lat = ~latitude, 
                   radius = 2, 
                   color = ~pal2(ethnicity),
                   group = "White",
                   popup = ~paste("<b>", ID, "</b>", "<br/>", date, "<br/>", age, "<br/>", sex, "<br/>", weapon, "<br/>", status)) %>%
  addCircleMarkers(data = asian,
                   lng = ~longitude, 
                   lat = ~latitude, 
                   radius = 2, 
                   color = ~pal2(ethnicity),
                   group = "Asian",
                   popup = ~paste("<b>", ID, "</b>", "<br/>", date, "<br/>", age, "<br/>", sex, "<br/>", weapon, "<br/>", status)) %>%
  addCircleMarkers(data = other,
                   lng = ~longitude, 
                   lat = ~latitude, 
                   radius = 2, 
                   color = ~pal2(ethnicity),
                   group = "Other",
                   popup = ~paste("<b>", ID, "</b>", "<br/>", date, "<br/>", age, "<br/>", sex, "<br/>", weapon, "<br/>", status)) %>%


  addLegend(position = "bottomright", 
            pal = pal2, 
            values = c("Black", "White", "Asian", "Other"),
            title = "Legend") %>%
  addLayersControl(
    overlayGroups = c("Black",
                      "White",
                      "Asian",
                      "Other"))

```
```
```

Column
-----------------------------------------------------------------------

### Gun homicide victims are younger on average than other victims (2008-2019)
```{r}
p19 <- ggplot (data = mmap8, aes(x = reorder(weapon, age, "mean"), y = age), na.rm = TRUE)+ geom_boxplot()+ geom_boxplot(outlier.shape = NA) + geom_jitter(alpha = 1/5) +
  ylab("Age of victim") +
  xlab("Cause of Death")+
  stat_summary(fun="mean", colour="red", geom="point",shape = 16, size = 4) +
  theme_minimal()
ggplotly(p19)


```


### Homicides involving guns are also less likely to get solved in the criminal justice system (2008-2020)

```{r}

p22 <- ggplot(mmap13, aes(x= reorder(weapon, percent), y=percent))+
geom_bar(stat="identity", fill="navyblue", color = "black")+
  coord_flip()+
  labs(x = "Weapon", y = "Percent Solved")+
  theme_minimal()
ggplotly(p22)


```



